---
title: "Google Capstone"
author: "Sarah Sabbah"
date: "14/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1. Document Description
This document shows steps taken to upload, clean and transform Cyclists datasets to prepare for analysis.

##First installing and loading the packages required to read, and clean data
```{r}
install.packages("tidyverse")
install.packages("janitor")

library(tidyverse)
library(janitor)

```

##then import 12 months data files from https://divvy-tripdata.s3.amazonaws.com/index.html
##while renaming each file into something more simple

```{r}
df1<-read.csv("202004-divvy-tripdata.csv")
df2<-read.csv("202005-divvy-tripdata.csv")
df3<-read.csv("202006-divvy-tripdata.csv")
df4<-read.csv("202007-divvy-tripdata.csv")
df5<-read.csv("202008-divvy-tripdata.csv")
df6<-read.csv("202009-divvy-tripdata.csv")
df7<-read.csv("202010-divvy-tripdata.csv")
df8<-read.csv("202011-divvy-tripdata.csv")
df9<-read.csv("202012-divvy-tripdata.csv")
df10<-read.csv("202101-divvy-tripdata.csv")
df11<-read.csv("202102-divvy-tripdata.csv")
df12<-read.csv("202103-divvy-tripdata.csv")
```

#now that we have all 12 data files imported, we want to combine them all into one dataframe using rbind, which stands for row bind
```{r}
bike_rides<-rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)
```

#now we do data cleaning by removing nulls in rows and columns
```{r}
bike_rides<-janitor::remove_empty(bike_rides,which=c("cols"))
bike_rides<-janitor::remove_empty(bike_rides,which=c("rows"))
```

##this shows the exact same number of columns and rows so therefore we didn't have duplicates to begin with
#we also do not want and na values
```{r}
bike_rides<-na.omit(bike_rides)
```

#okay so this did remove some rows, we went down from 3489748 rows to 3354514 rows
#now we ensure that data types are correct, we want to change chr to a date time format
```{r}
bike_rides$started_at<-as.POSIXct(bike_rides$started_at,"%Y-%m-%d %H:%M:%S")
bike_rides$ended_at<-as.POSIXct(bike_rides$ended_at,"%Y-%m-%d %H:%M:%S")
```
#we also want to make a new column with ride lenght and ensure that ride duration is entered as numeric
#load dplyr so %>% works

```{r}
install.packages("dplyr")
library(dplyr)
```

```{r}
bike_rides<-bike_rides %>%
  mutate(ride_length=as.numeric(bike_rides$ended_at-bike_rides$started_at)/60)
```

#works!you can check now that your table has 14 variables(columns)

#now lets looks at the average of ride times
```{r}
summary(bike_rides$ride_length)
```

#this gives you nonsensical numbers, but looks like the average(mean)is ride is 24.55

#lets make sure no ride times under 0 are entered

```{r}
bike_rides<-bike_rides %>%
  filter(!(ride_length <0))
```
#this has eliminated rows, we now have 3351199 obs
#lets get an average again
```{r}
summary(bike_rides$ride_length)

```
#so this changes our average to 27.79
#now we want to create a column for year, month and day of the week, date and time of day
```{r}
bike_rides$year<-format(
  bike_rides$started_at,
  "%Y"
)
```
```{r}
bike_rides$month<-format(
  bike_rides$started_at,
  "%m"
)
```
```{r}
bike_rides$weekday<-format(
  bike_rides$started_at,
  "%A"
)
```
```{r}
bike_rides$YMD<-format(
  bike_rides$started_at,
  "%Y-%m-%d"
)
```
```{r}
bike_rides$ToD<-format(
  bike_rides$started_at,
  "%H:%M:%S"
)
```
to continue cleaning, we want to make sure there isn't any duplicates
```{r}
ride_id_check<-bike_rides%>%
  count(ride_id)%>%
  filter(n > 1)
```
#this takes a couple of minutes, but no duplicates are found
#at this point, this dataset is cleaned, so save the dataframe, this also takes a few minutes (we have 19 columns and 3351199 rows)
```{r}
bike_rides%>%
  write.csv("bike_rides_cleaned.csv")
```
#now i want to understand what and how many types of bikes there is 
```{r}
unique(bike_rides$rideable_type)
```

#this returns 3 options:docked, electric and classic
#i want to see how many stations there are
```{r}
station_name_check<-bike_rides %>%
  group_by(start_station_name) %>%
  count(start_station_name)
```
#wow, there is 706 stations

#now i want to know what stations where used at each month, so my data starts in april of 2020 to march of 2021
```{r}
April_2020_filter<-bike_rides %>%
  filter(
    month =="04"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

# great so i'm going to do this for all months
```{r}
May_2020_filter<-bike_rides%>%
  filter(
    month =="05"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
June_2020_filter<-bike_rides%>%
  filter(
    month =="06"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
July_2020_filter<-bike_rides%>%
  filter(
    month =="07"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
August_2020_filter<-bike_rides%>%
  filter(
    month =="08"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
September_2020_filter<-bike_rides%>%
  filter(
    month =="09"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
October_2020_filter<-bike_rides%>%
  filter(
    month =="10"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
November_2020_filter<-bike_rides%>%
  filter(
    month =="11"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
December_2020_filter<-bike_rides%>%
  filter(
    month =="12"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
January_2021_filter<-bike_rides%>%
  filter(
    month =="01"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```


```{r}
February_2021_filter<-bike_rides%>%
  filter(
    month =="02"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```

```{r}
March_2021_filter<-bike_rides%>%
  filter(
    month =="03"
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(start_station_name)
```
#okay so now I've made dataframes with each month and also how many times the station was used
#now we want to check if a type of bike was added in a specific month, knowing this will give context to trends
```{r}
rideable_type<-bike_rides %>% 
  group_by(
    month,
    year
  ) %>% 
  select(
  rideable_type,
  month,
  year
) %>% 
count(
  rideable_type
)
```
```{r}
library(dplyr)
```
#this gives us a dataframe to view the bikes rides used most in what month as well as when a new type was introduced.

```{r}
rideable_type %>% 
  group_by(rideable_type) %>% 
  filter(year==min(year)) %>% 
  filter(month==min(month))
```
#this shows that classic bikes where added in December of 2020, and electric bikes added in July 2020, prior to that, the only option for riders were docked bikes

#now we want to assess how many times each stations was used in each month

```{r}
station_name_check$April_2020<-as.integer(station_name_check$start_station_name
                                          %in% April_2020_filter$start_station_name)

station_name_check$May_2020<-as.integer(station_name_check$start_station_name
                                          %in% May_2020_filter$start_station_name)

station_name_check$June_2020<-as.integer(station_name_check$start_station_name
                                          %in% June_2020_filter$start_station_name)

station_name_check$July_2020<-as.integer(station_name_check$start_station_name
                                          %in% July_2020_filter$start_station_name)

station_name_check$August_2020<-as.integer(station_name_check$start_station_name
                                          %in% August_2020_filter$start_station_name)

station_name_check$September_2020<-as.integer(station_name_check$start_station_name
                                          %in% September_2020_filter$start_station_name)

station_name_check$October_2020<-as.integer(station_name_check$start_station_name
                                          %in% October_2020_filter$start_station_name)

station_name_check$November_2020<-as.integer(station_name_check$start_station_name
                                          %in% November_2020_filter$start_station_name)

station_name_check$December_2020<-as.integer(station_name_check$start_station_name
                                          %in% December_2020_filter$start_station_name)

station_name_check$January_2021<-as.integer(station_name_check$start_station_name
                                          %in% January_2021_filter$start_station_name)

station_name_check$February_2021<-as.integer(station_name_check$start_station_name
                                          %in% February_2021_filter$start_station_name)

station_name_check$March_2021<-as.integer(station_name_check$start_station_name
                                          %in% March_2021_filter$start_station_name)

```

add total/sum column

```{r}
station_name_check$count<-rowSums(station_name_check[,3:14])

```

#data is now cleaned and ready for analysis

a few questions we will answer are: 
what are the most popular stations
most popular times of the year
most popular day of the week
most popular time of day
bike type preferences

```{r}
library(ggplot2)
```

```{r}
bike_rides %>% 
  group_by(member_casual,weekday) %>% 
  summarise(average_duration=mean(ride_length)) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x=weekday,y=average_duration,
              fill=member_casual))+
  geom_col(position = "dodge")+
  scale_y_continuous(labels=function(x)format(x, scientific=FALSE))+
  labs(title="Average Ride Duration(Minutes) by User Type")+
  ylab("Average Duration in Minutes")+
  xlab("Day of Week")
#ride duration

```
```{r}
bike_rides %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides=n()) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x=weekday,y=number_of_rides,
             fill=member_casual))+
  geom_col(position = "dodge")+
  scale_y_continuous(labels=function(x)format(x, scientific=FALSE))+
  labs(title = "Average Rides by User Type")+
  ylab("Number of Rides")+
  xlab("Day of Week")+
  labs(fill="User Type")
#rides by Weekday
```

Rides by Month
```{r}
bike_rides %>% 
  group_by(member_casual,month) %>% 
  summarise(number_of_rides=n()) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x=month,y=number_of_rides,
             fill=member_casual))+
  geom_col(position = "dodge")+
  scale_y_continuous(labels=function(x)format
                     (x,scientific=FALSE))+
  labs(title="Total Month Rides by User Type")+
  ylab("Average Number of Rides")+
  xlab("Month")+
  labs(fill="User Type")
  

```
Bike Type Preference 
```{r}
bike_rides %>%
  group_by(member_casual,month,rideable_type) %>% 
  ggplot(aes(x=month, y=rideable_type,
             fill=member_casual))



```


